Cognate beginnings to bilingual lexical acquisition

### 2023-07-23 manuscript/manuscript.tex
Simplify dependencies

source of this study. GGC was supported by the a FPI 2020 research
contract (PRE2019-088165) in a project funded by the Spanish Ministry of
Science, Innovation, and Universities (PGC2018-101831-B-I00). DAV was
supported by the European Union's Horizon 2023 research and innovation
program under Marie Skłodowska--Curie Grant (765556) and a postdoctoral

@@

(UIDB/00214/2020). NSG was supported by Academia 2019 award from the
Catalan Institution for Research and Advanced Studies (ICREA). We are
grateful to Chiara Santolin, Ege E. Özer, Alicia Franco-Martínez,
Cristina Rodríguez-Prada, and the rest of the Speech Acquisition and
Perception research group for their helpful feedback. We thank Xavier
Mayoral, Silvia Blanch, and Cristina Cuadrado for their technical
support, and Cristina Dominguez and Katia Pistrin for their efforts in
recruiting infants. We also thank all families and infants who
participated in the experiments.}}

@@

Bilingual infants show equivalent developmental trajectories of lexical
acquisition compared to their monolingual peers. This is remarkable,
given the increased complexity of their linguistic input. Recent studies
suggested that bilingual vocabulary growth is boosted by the number of
cognates shared by the pair of languages being learned, and that this
facilitation effect is driven by a stronger parallel activation of
cognates during linguistic exposure, compared to non-cognates. The
mechanisms behind this facilitation are still unclear. In this study, we
capitalise on accumulator models of language acquisition to propose an
account of bilingual lexical acquisition in which parallel activation
increases the rate at which children accumulate learning instances for
words in both languages, even in fully monolingual situations. Under
this hypothesis, we predicted a stronger cognate facilitation for words
to which children were exposed less frequently (low-exposure words), as
they are co-activated by their translation more often than high-exposure
words do. We developed an extensive online vocabulary checklist, the
Barcelona Vocabulary questionnaire (BVQ) to collect vocabulary data from
366 Catalan-Spanish bilingual toddlers aged 12 to 32 months. We then
used Bayesian explanatory item response theory to model the acquisition
trajectories of 302 Catalan/Spanish translation equivalents. We found an
interaction between exposure and cognateness, which pointed to
cognateness facilitating the aquisition of low-exposure words, but not
of mean exposure or high-exposure words. Overall, our findings suggest
that cognateness plays a key role in bilingual lexical acquisition, and

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

The foundations of word learning are in place early in age. At six
months, infants start directing their gaze to some objects when hearing
their labels (\protect\hyperlink{ref-bergelson2012months}{Bergelson \&
Swingley, 2012}, \protect\hyperlink{ref-bergelson2015early}{2015};
\protect\hyperlink{ref-tincoff1999beginnings}{Tincoff \& Jusczyk,
1999}), and shortly after caregivers start reporting some words as
acquired by their infant in vocabulary checklists (e.g.,
\protect\hyperlink{ref-fenson2007macarthurbates}{Fenson et al., 2007};
\protect\hyperlink{ref-samuelson2021precision}{Samuelson, 2021}). Most
research on early word acquisition relies extensively on data from
monolingual children, and is oblivious to the fact that a substantial
proportion of the world population acquires more than one language from
early ages (\protect\hyperlink{ref-grosjean2021extent}{Grosjean, 2021}).
Previous work on bilingual vocabulary acquisition pointed to bilingual
toddlers knowing, on average, less words in each of their languages than
their monolinguals peers, and to both groups knowing a similar number of
words---if not more words---when the bilinguals' two languages are
pooled together. Hoff et al.
(\protect\hyperlink{ref-hoff2012dual}{2012}) found that English-Spanish
bilingual toddlers in South Florida (United States) knew less words in
English than monolinguals did, but both groups knew a similar total
amount of words when both English and Spanish vocabularies were counted
together. Other studies have provided converging evidence that
bilinguals know a similar or even larger number of words than
monolinguals when the two languages are aggregated
(\protect\hyperlink{ref-gonzalez-barrero2020bilingual}{Gonzalez-Barrero
et al., 2020}; \protect\hyperlink{ref-oller2002language}{Oller \&
Eilers, 2002}; \protect\hyperlink{ref-patterson2004comparing}{Patterson,
2004}; \protect\hyperlink{ref-patterson2004bilingual}{Patterson \&
Pearson, 2004}; \protect\hyperlink{ref-pearson1993lexical}{Pearson et
al., 1993}; \protect\hyperlink{ref-pearson1994patterns}{Pearson \&
Fernández, 1994}; \protect\hyperlink{ref-petitto2001bilingual}{Petitto
et al., 2001}; \protect\hyperlink{ref-smithson2014bilingualism}{Smithson
et al., 2014}). A more detailed analysis of the words in bilinguals'
lexicon show some interesting patterns.

One important observation of studies on bilinguals' early vocabulary
acquisition is that cognate words are easier to acquire than non-cognate
words. Cognate words are translation equivalents that are phonologically
similar (or share some type of form-similarity). For instance, the
Spanish translation equivalent of \emph{cat} is \emph{gato}, a cognate
word; the translation equivalent of \emph{dog} is \emph{perro}, a
non-cognate word. The differences in the percentages of cognate and
non-cognate words between two languages is related to historic reasons:
languages typologically close (like Dutch and English or Italian and
Spanish) share more cognates than languages typologically distant (like
English and Chinese, or Urdu and Spanish). The conclusion that cognate
words are easier to learn is based on two types of evidence: studies
investigating vocabulary sizes in children learning language pairs with
different percentages of cognates (that is, differing in their
typological distance) and studies comparing the number of cognate and
non-cognate words children know in a specific language pair.

Floccia et al. (\protect\hyperlink{ref-floccia2018introduction}{2018})
published an impressive study comparing vocabularies of children
learning several language pairs differing in their percentage of
cognates. The authors collected vocabulary data on word comprehension
and production from 372 24-month-old bilingual toddlers living in the
United Kingdom who were learning English and an additional language. The
additional language was one of 13 typologically diverse languages:
Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi/Urdu,
Italian, Mandarin Chinese, Polish, Portuguese, Spanish, and Welsh. The
authors calculated the average phonological overlap between the words in
each of these additional languages and their translation equivalents in
English, which was taken as a proxy of the degree of cognateness between
each pair of languages. Floccia and co-workers reported an increase in
vocabulary size in the additional language (i.e., not English)
associated with an increase in the average phonological similarity
between the translation equivalents of each language pair. For example,
English-Dutch bilinguals (languages with a high phonological overlap),
were able to produce more Dutch words than English-Mandarin bilinguals
(languages with a low phonological overlap) were able to produce in
Mandarin. Blom et al.
(\protect\hyperlink{ref-blom2020crosslanguage}{2020}), Bosma et al.
(\protect\hyperlink{ref-bosma2019longitudinal}{2019}), and Gampe et al.
(\protect\hyperlink{ref-gampe2021does}{2021}) reported similar results,
providing converging evidence of a facilitatory effect of a lower
language distance (i.e., higher degre of cognateness) on vocabulary
size.

A second set of studies concluded that cognates are overrepresented in
bilinguals' early lexicon. Bosch \& Ramon-Casas
(\protect\hyperlink{ref-bosch2014first}{2014}) collected parental
reports of expressive vocabulary from 48 Catalan-Spanish bilinguals aged
18 months and found that cognates represented a larger proportion of
participant's vocabulary than non-cognates. Schelletter
(\protect\hyperlink{ref-schelletter2002effect}{2002}) provided
converging evidence from a longitudinal single-case study, in which an
English-German bilingual child produced cognates earlier than
non-cognates, on average. Mitchell et al.
(\protect\hyperlink{ref-mitchell2022cognates}{2022}) addressed this
issue in a larger sample longitudinal study. These authors collected
expressive vocabulary data of 47 16- to 30-month-old French-English
bilinguals living in Canada, in both languages. They created two lists
of translation equivalents: one made of 131 cognates, and one made of
406 non-cognates. The proportion of words that children were reportedly
able to produce was higher in the cognate lists than in the non-cognate
list across ages, even when both lists were matched by semantic category
(furniture, animals, food were similarity represented in both lists) and
age-of-acquisition norms (an index of word difficulty). Taken together,
the results of these two lines of research provide a strong support to
the hypothesis that phonological similarity (as reflected in
cognateness) plays a facilitation role in bilingual word acquisition.

@@

We propose an account in which a learning instance for a word may also
represent a learning instance for its translation equivalent, to the
extent that such translation equivalent is co-activated\footnote{We use
  the term \emph{learning instance} in the fashion of accumulator models
  of language acquisition; as an exposure to a word-form that
  constitutes an opportunity for the child to accumulate information
  about the word. We do not make strong assumptions about whether a
  learning instance is a discrete or a continuous unit of accumulation
  of information. We rather consider that a learning instance of a word
  is so to the degree of the strength of activation of its phonological
  representation. This activation may result from the infant being
  exposed to the word-form, or the result of activation spreading
  through phonological or semantic links across lexical representation,
  as in the case of parallel activation.}. The strength of this
co-activation is proportional to the phonological similarity between the
two translation equivalents; given that cognates share higher
phonological similarity than non-cognates, the former should be
co-activated more frequently than the latter. This should lead to a
faster accumulation of learning instances for cognates, compared to
non-cognates. Parallel activation would allow bilingual children to
accumulate learning instances for words in both languages even during
fully monolingual situations, but the impact of this mechanism would be
asymmetric across languages: words from the lower-exposure language
(i.e., the language to which the infants are exposed to more often)
would receive additional activation through parallel activation more
frequently than words from the higher-exposure language. Therefore, the
acquisition of words from the lower-exposure language would benefit more
strongly from their cognate status than words from the higher-exposure
language.

@@

We collected 436 responses to the questionnaire from 366 distinct
children from the Metropolitan Area of Barcelona between the 30th of
March, 2020 and the 31th of October, 2022: 312 of those participants
participated once, 42 twice, 8 three times, and 4 four times. Recurrent
participants provided responses with a minimum of 25 days between
responses, and a maximum of 527, and were always allocated to the same
questionnaire list (A, B, C, or D). Participants were part of the
database of the Laboratori de Recerca en Infància (Universitat Pompeu
Fabra) and were contacted by e-mail or phone if their child was aged
between 12 and 32 months, and had not been reported to be exposed more
than 10\% of the time to a language other than Spanish or Catalan (see
Table~\ref{tbl-participants} for a more detailed description of the
sample). In total, 70 participants (16.06\%) participants were reported
to be exposed to a third language other than Catalan and Spanish. All
families provided informed consent before participating. Upon consent,
families were sent a link to the questionnaire via e-mail, which they
filled from a computer, laptop, or mobile device. Filling the
questionnaire took 30 minutes approximately. After completion, families
were rewarded with a token of appreciation.

@@

\centering
\begin{tabular}{lcccccc}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{6}{c}{Age (months)} \\
\cmidrule(l{3pt}r{3pt}){2-7}
Catalan exposure & {}[10-14] & (14, 18] & (18, 22] & (22-26] & (26-30] & (30-34]\\
\midrule
75-100\% & 18 & 23 & 36 & 38 & 20 & 7\\
50-75\% & 8 & 13 & 30 & 41 & 18 & 1\\
25-50\% & 10 & 17 & 45 & 29 & 17 & 0\\
0-25\% & 7 & 11 & 21 & 17 & 8 & 1\\
\midrule
N & 43 & 64 & 132 & 125 & 63 & 9\\
\bottomrule
\end{tabular}
\end{table}

We used the highest self-reported educational attainment of parents or
caregivers as a proxy of participants' socio-economic status (SES). This
information was provided by each parent or caregiver by selecting one of
six possible alternatives in line with the current educational system in
Spain: \emph{sense escolaritzar/sin escolarizar} {[}no education{]},
\emph{educació primària/educación primaria} {[}primary school{]},
\emph{educació secundària/educación secundaria} {[}secondary school{]},
\emph{batxillerat/bachillerato} {[}complementary studies/high school{]},
\emph{cicles formatius/ciclos formativos} {[}vocational training{]}, and
\emph{educació universitària/educación universitaria} {[}university
degree{]}. Most families reported university studies (356, 82\%),
followed by families were the highest educational attainment were
vocational studies (59, 14\%), secondary education (8, 2\%),
complementary studies (6, 1\%), primary education (1, \textless1\%), and
no formal education (2, \textless1\%).

@@

We collected data for 1,590 words. We restricted the analyses to
responses to nouns (628 items corresponding to other grammatical classes
were excluded, see \protect\hyperlink{ref-fourtassi2020growth}{Fourtassi
et al., 2020} for a similar approach). We then excluded items with
missing lexical frequency scores (\emph{n} = 269), items that included
more than one lemma (e.g., \emph{mono/mico} {[}monkey{]}, \emph{n} =
48), multi-word items or phrases (e.g., \emph{barrita de cereales}
{[}cereal bar{]}, \emph{n} = 9). Finally, we removed items without a
translation in the other language (\emph{n} = 32). This resulted in a
final list of 604 items, corresponding to 302 Catalan words and their
302 Spanish translations (302 translation equivalents). After collecting
participants' responses, the final dataset consisted of 138,078
observations, each corresponding to a single response of one participant
to one item. Each translation equivalent received a median of 234
responses (\emph{Min} = 106, \emph{Max} = 872) from participants, both
languages pooled together.

@@

We developed the \emph{Exposure} predictor to account for the fact that
bilinguals' exposure to a given word-form is not only a function of the
word-forms lexical frequency, but also of the quantitative input they
receive from the language such word-form belongs to, we expressed
lexical frequencies as the product between both variables. First, we
extracted the child-directed lexical frequency of each word from the
CHILDES database
(\protect\hyperlink{ref-macwhinney2000childes}{MacWhinney, 2000};
\protect\hyperlink{ref-sanchez2019childesdb}{Sanchez et al., 2019}).
Using the corresponding lexical frequencies directly from Catalan and
Spanish was not possible due to the low number of Catalan participants
and tokens available in their corresponding CHILDES corpora, so they
were extracted from the English corpora instead. We then mapped the
lexical frequencies of the English words to their Catalan and Spanish
translations (see \protect\hyperlink{ref-fourtassi2020growth}{Fourtassi
et al., 2020} for a similar approach), and transformed them to Zipf
scores (\protect\hyperlink{ref-vanheuven2014subtlexuk}{Van Heuven et
al., 2014}; \protect\hyperlink{ref-zipf1949human}{Zipf, 1949}). We then
multiplied the resulting lexical frequencies by the reported degree of
exposure of the child to Catalan or Spanish. For instance, for a child
whose degree of exposure is 80\% for Catalan and 20\% for Spanish, the
expected \emph{Exposure} score to the Catalan word \emph{cotxe}
{[}\emph{car}{]}---with a lexical frequency of 6.33---would be 5.06,
while that of its translation to Spanish \emph{coche} would be 1.27.

We defined \emph{Cognateness} as the phonological similarity between
each word-form and its translation. For each translation equivalent, we
used the \texttt{stringdist}
(\protect\hyperlink{ref-vanderloo2014stringdist}{Loo, 2014}) R package
to calculate the Levenshtein distance between the Catalan and the
Spanish phonological transcriptions of the word-forms. The Levenshtein
distance measures the number of editions (insertions, deletions, or
substitutions) that one string of phonemes/characters must go through to
become identical to the other
(\protect\hyperlink{ref-levenshtein1966binary}{Levenshtein, 1966}). We
divided the Levenshtein distance of each translation equivalent by the
length of the longest word-form to correct for word length (longer
strings are likely to show a larger number of mismatches). Finally, we
subtracted the result from one so that it could be interpreted in terms
of phonological similarity, instead of phonological distance. This led
to a distance metric that ranged from zero to one, where zero indicates
that both word-forms are completely different (e.g.,
/\textipa{"taw.l5}-\textipa{"me.sa}/, \emph{table}), and one indicates
that the two word-forms are identical (e.g.,
/\textipa{"mar}-\textipa{"mar}/, \emph{sea}) (see
\protect\hyperlink{ref-floccia2018introduction}{Floccia et al., 2018};
\protect\hyperlink{ref-fourtassi2020growth}{Fourtassi et al., 2020};
\protect\hyperlink{ref-heeringa2003norwegian}{Heeringa \& Gooskens,
2003}; \protect\hyperlink{ref-laing2022phonological}{Laing, 2022} for
similar approaches;
\protect\hyperlink{ref-schepens2012distributions}{Schepens et al.,
2012}).

@@

We assessed the practical relevance of the estimated regression
coefficients of the model model following Kruschke \& Liddell
(\protect\hyperlink{ref-kruschke2018bayesian}{2018}). First, we
specified a region of practical equivalence (ROPE) from -2.5\% to
+2.5\%, in the probability scale. This region indicates the range of
values that we considered equivalent to zero. We then summarised the
posterior distribution of each regression coefficient with the 95\%
highest density interval (HDI). This interval contains the true value of
this coefficient with 95\% probability, given the data. Finally, we
calculated the proportion of posterior samples in the 95\% HDI that fell
into the ROPE, noted as \(p(\text{ROPE})\), which indicates the
probability that the true value of the regression coefficient falls into
the ROPE (and therefore should be considered equivalent to zero). For
example, \(p(\text{ROPE})=.80\) indicates that, given our data, there is
a 80\% probability that the true value of the coefficient falls within
the ROPE, and can therefore be considered equivalent to zero.

>>>>

source of this study. GGC was supported by a FPI 2020 research contract
(PRE2019-088165) in a project funded by the Spanish Ministry of Science,
Innovation, and Universities (PGC2018-101831-B-I00). DAV was supported
by the European Union's Horizon 2023 research and innovation program
under Marie Skłodowska--Curie Grant (765556) and a postdoctoral

@@

(UIDB/00214/2020). IC was supported by the Investigo program funded by
the European Union's NextGenerationEU (NGEU) recovery plan. NSG was
supported by Academia 2019 award from the Catalan Institution for
Research and Advanced Studies (ICREA). We are grateful to Chiara
Santolin, Ege E. Özer, and the rest of the Speech Acquisition and
Perception research group, and to Alicia Franco-Martínez and Cristina
Rodríguez-Prada, for their helpful feedback. We thank Xavier Mayoral,
Silvia Blanch, and Cristina Cuadrado for their technical support, and
Cristina Dominguez and Katia Pistrin for their efforts in recruiting
infants. We also thank all families and infants who participated in the
experiments.}}

@@

Bilingual infants' developmental trajectories of lexical acquisition are
equivalent to those shown by their monolingual peers. This is
remarkable, given the increased complexity of their linguistic input.
Recent studies suggested that bilingual vocabulary growth is boosted by
the number of cognates shared by the pair of languages being learned,
and that this facilitation effect is driven by a stronger parallel
activation of cognates during linguistic exposure, compared to
non-cognates. The mechanisms behind this facilitation are still unclear.
In this study, we capitalised on accumulator models of language
acquisition to propose an account of bilingual lexical acquisition in
which parallel activation increases the rate at which children
accumulate learning instances for words in both languages, even in fully
monolingual situations. Under this hypothesis, we predicted a stronger
cognate facilitation for words to which children were exposed less
frequently (low-exposure words), as they are co-activated by their
translation more often than high-exposure words do. We developed an
extensive online vocabulary checklist, the Barcelona Vocabulary
questionnaire (BVQ), to collect vocabulary data from 366 Catalan-Spanish
bilingual toddlers aged 12 to 32 months. We then used Bayesian
explanatory item response theory to model the acquisition trajectories
of 302 Catalan/Spanish translation equivalents. We found an interaction
between exposure and cognateness, which pointed to cognateness
facilitating the aquisition of low-exposure words, but not of mean
exposure or high-exposure words. Overall, our findings suggest that
cognateness plays a key role in bilingual lexical acquisition, and

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, interior hidden, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners]}{\end{tcolorbox}}\fi

@@

The foundations of word learning are in place early in age. At six
months, infants start directing their gaze to some objects when hearing
their labels (\protect\hyperlink{ref-bergelson2012months}{Bergelson \&
Swingley, 2012}, \protect\hyperlink{ref-bergelson2015early}{2015};
\protect\hyperlink{ref-tincoff1999beginnings}{Tincoff \& Jusczyk,
1999}), and shortly after caregivers start reporting some words as
acquired by their infant in vocabulary checklists (e.g.,
\protect\hyperlink{ref-fenson2007macarthurbates}{Fenson et al., 2007};
\protect\hyperlink{ref-samuelson2021precision}{Samuelson, 2021}). Most
research on early word acquisition relies extensively on data from
monolingual children, and is oblivious to the fact that a substantial
proportion of the world population acquires more than one language from
early ages (\protect\hyperlink{ref-grosjean2021extent}{Grosjean, 2021}).
Previous work on bilingual vocabulary acquisition pointed to bilingual
toddlers knowing, on average, less words in each of their languages than
their monolinguals peers, and to both groups knowing a similar number of
words---if not more words---when the bilinguals' two languages are
pooled together. Hoff et al.
(\protect\hyperlink{ref-hoff2012dual}{2012}) found that English-Spanish
bilingual toddlers in South Florida (United States) knew less words in
English than monolinguals did, but both groups knew a similar total
amount of words when both English and Spanish vocabularies were counted
together. Other studies have provided converging evidence that
bilinguals know a similar or even larger number of words than
monolinguals when the two languages are aggregated
(\protect\hyperlink{ref-gonzalez-barrero2020bilingual}{Gonzalez-Barrero
et al., 2020}; \protect\hyperlink{ref-oller2002language}{Oller \&
Eilers, 2002}; \protect\hyperlink{ref-patterson2004comparing}{Patterson,
2004}; \protect\hyperlink{ref-patterson2004bilingual}{Patterson \&
Pearson, 2004}; \protect\hyperlink{ref-pearson1993lexical}{Pearson et
al., 1993}; \protect\hyperlink{ref-pearson1994patterns}{Pearson \&
Fernández, 1994}; \protect\hyperlink{ref-petitto2001bilingual}{Petitto
et al., 2001}; \protect\hyperlink{ref-smithson2014bilingualism}{Smithson
et al., 2014}). A more detailed analysis of the words in bilinguals'
lexicon shows some interesting patterns.

One important observation of studies on bilinguals' early vocabulary
acquisition is that cognate words are easier to acquire than non-cognate
words. Cognate words are translation equivalents that are phonologically
similar (or share some type of form-similarity). For instance, the
Spanish translation equivalent of \emph{cat} is \emph{gato}, a cognate
word; the translation equivalent of \emph{dog} is \emph{perro}, a
non-cognate word. The differences in the percentages of cognate and
non-cognate words between two languages is related to historical
reasons: languages typologically close (like Dutch and English or
Italian and Spanish) share more cognates than languages typologically
distant (like English and Chinese, or Urdu and Spanish). The conclusion
that cognate words are easier to learn is based on two types of
evidence: studies investigating vocabulary sizes in children learning
language pairs with different percentages of cognates (that is,
differing in their typological distance) and studies comparing the
number of cognate and non-cognate words children know in a specific
language pair.

Floccia et al. (\protect\hyperlink{ref-floccia2018introduction}{2018})
published an impressive study comparing vocabularies of children
learning several language pairs differing in their percentage of
cognates. The authors collected vocabulary data on word comprehension
and production from 372 24-month-old bilingual toddlers living in the
United Kingdom who were learning English and an additional language. The
additional language was one of 13 typologically diverse languages:
Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi/Urdu,
Italian, Mandarin Chinese, Polish, Portuguese, Spanish, and Welsh. The
authors calculated the average lexical overlap between the words in each
of these additional languages and their translation equivalents in
English, which was taken as a proxy of the degree of cognateness between
each pair of languages. Floccia and co-workers reported an increase in
vocabulary size in the additional language (i.e., not English)
associated with an increase in the average phonological similarity
between the translation equivalents of each language pair. For example,
English-Dutch bilinguals (languages with a high phonological overlap),
were able to produce more Dutch words than English-Mandarin bilinguals
(languages with a low phonological overlap) were able to produce in
Mandarin. Blom et al.
(\protect\hyperlink{ref-blom2020crosslanguage}{2020}), Bosma et al.
(\protect\hyperlink{ref-bosma2019longitudinal}{2019}), and Gampe et al.
(\protect\hyperlink{ref-gampe2021does}{2021}) reported similar results,
providing converging evidence of a facilitatory effect of a lower
language distance (i.e., higher degre of cognateness) on vocabulary
size.

A second set of studies suggested that cognates are overrepresented in
bilinguals' early lexicon. Bosch \& Ramon-Casas
(\protect\hyperlink{ref-bosch2014first}{2014}) collected parental
reports of expressive vocabulary from 48 Catalan-Spanish bilinguals aged
18 months and found that cognates represented a larger proportion of
participant's vocabulary than non-cognates. Schelletter
(\protect\hyperlink{ref-schelletter2002effect}{2002}) provided
converging evidence from a longitudinal single-case study, in which an
English-German bilingual child produced cognates earlier than
non-cognates, on average. But the high proportion of cognates in the
vocabulary of the participants in these two studies may not necessarily
evidence of a facilitation effect of cognateness, but rather of simply
the high proportion of cognates present in the pair of languages being
learned. For instance, if two given languages share a high proportion of
cognates like \(\sim 70\)\%, the vocabulary contents of children
learning both languages should, in principle, approximate such
proportion of cognates, even in the absence of a cognateness
facilitation effect. More recently, Mitchell et al.
(\protect\hyperlink{ref-mitchell2022cognates}{2022}) addressed this
issue in a longitudinal study with a larger sample. The authors
collected expressive vocabulary data of 47 16- to 30-month-old
French-English bilinguals living in Canada, in both languages. They
created two lists of translation equivalents: one made of 131 cognates,
and one made of 406 non-cognates. The proportion of words that children
were reportedly able to produce was higher in the cognate lists than in
the non-cognate list across ages. Critically, this difference persisted
after both lists were matched in size, controlling their semantic
category (i.e., furniture, animals, food were similarity represented in
both lists) and age-of-acquisition norms (an index of word difficulty).
Taken together, the results of these two lines of research have been
considered as providing support to the hypothesis that phonological
similarity (as reflected in cognateness) plays a facilitation role in
bilingual word acquisition.

@@

We propose an account in which a learning instance for a word may also
represent a learning instance for its translation equivalent, to the
extent that such translation equivalent is co-activated\footnote{We use
  the term \emph{learning instance} in the fashion of accumulator models
  of language acquisition; as an exposure to a word-form that
  constitutes an opportunity for the child to accumulate information
  about the word. We do not make strong assumptions about whether a
  learning instance is a discrete or a continuous unit of accumulation
  of information. We rather consider that a learning instance of a word
  is so to the degree of the strength of activation of its phonological
  representation. This activation may result from the infant being
  exposed to the actual word-form, or the result of activation spreading
  through phonological or semantic links across lexical representation,
  as in the case of parallel activation.}. The strength of this
co-activation is proportional to the phonological similarity between the
two translation equivalents; given that cognates share higher
phonological similarity than non-cognates, the former should be
co-activated more strongly than the latter. This should lead to a faster
accumulation of learning instances for cognates, compared to
non-cognates. Parallel activation would allow bilingual children to
accumulate learning instances for words in both languages even during
fully monolingual situations, but the impact of this mechanism would be
asymmetric across languages: words from the lower-exposure language
would receive stronger activation from words in the higher-exposure
language than vice versa. Therefore, the acquisition of words from the
lower-exposure language would benefit more strongly from their cognate
status than words from the higher-exposure language. This asymmetric
cross-language activation would be consistent with previous reports of
larger priming effects from the dominant to the non-dominant language,

@@

We collected 440 responses to the questionnaire from 369 distinct
children from the Metropolitan Area of Barcelona between the 30th of
marzo, 2020 and the 31th of octubre, 2022: 314 of those participants
participated once, 43 twice, 8 three times, and 4 four times. Recurrent
participants provided responses with a minimum of 25 days between
responses, and a maximum of 527, and were always allocated to the same
questionnaire list (A, B, C, or D). Participants were part of the
database of the Laboratori de Recerca en Infància (Universitat Pompeu
Fabra) and were contacted by e-mail or phone if their child was aged
between 12 and 32 months, and had not been reported to be exposed more
than 10\% of the time to a language other than Spanish or Catalan (see
Table~\ref{tbl-participants} for a more detailed description of the
sample). In total, 70 participants (15.91\%) participants were reported
to be exposed to a third language other than Catalan and Spanish. All
families provided informed consent before participating. Upon consent,
families were sent a link to the questionnaire via e-mail, which they
filled from a computer, laptop, or mobile device. Filling the
questionnaire took 30 minutes approximately. After completion, families
were rewarded with a token of appreciation.

@@

\centering
\begin{tabular}{lcccccc}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{6}{c}{Age (months)} \\
\cmidrule(l{3pt}r{3pt}){2-7}
Catalan exposure & {}[10-14] & (14, 18] & (18, 22] & (22-26] & (26-30] & (30-34]\\
\midrule
75-100\% & 18 & 23 & 37 & 38 & 20 & 7\\
50-75\% & 8 & 13 & 30 & 41 & 18 & 1\\
25-50\% & 10 & 18 & 46 & 30 & 17 & 0\\
0-25\% & 7 & 11 & 21 & 17 & 8 & 1\\
\midrule
N & 43 & 65 & 134 & 126 & 63 & 9\\
\bottomrule
\end{tabular}
\end{table}

We used the highest self-reported educational attainment of parents or
caregivers as a proxy of participants' socio-economic status (SES). This
information was provided by each parent or caregiver by selecting one of
six possible alternatives in line with the current educational system in
Spain: \emph{sense escolaritzar/sin escolarizar} {[}no education{]},
\emph{educació primària/educación primaria} {[}primary school{]},
\emph{educació secundària/educación secundaria} {[}secondary school{]},
\emph{batxillerat/bachillerato} {[}complementary studies/high school{]},
\emph{cicles formatius/ciclos formativos} {[}vocational training{]}, and
\emph{educació universitària/educación universitaria} {[}university
degree{]}. Most families reported university studies (358, 82\%),
followed by families were the highest educational attainment were
vocational studies (61, 14\%), secondary education (8, 2\%),
complementary studies (6, 1\%), primary education (1, \textless1\%), and
no formal education (2, \textless1\%).

@@

We collected data for 1,590 words. We restricted the analyses to
responses to nouns (628 items corresponding to other grammatical classes
were excluded, see \protect\hyperlink{ref-fourtassi2020growth}{Fourtassi
et al., 2020} for a similar approach). We then excluded items with
missing lexical frequency scores (\emph{n} = 269), items that included
more than one lemma (e.g., \emph{mono/mico} {[}monkey{]}, \emph{n} =
48), multi-word items or phrases (e.g., \emph{barrita de cereales}
{[}cereal bar{]}, \emph{n} = 9). Finally, we removed items without a
translation in the other language (\emph{n} = 32). This resulted in a
final list of 604 items, corresponding to 302 Catalan words and their
302 Spanish translations (302 translation equivalents). After collecting
participants' responses, the final dataset consisted of 139,326
observations, each corresponding to a single response of one participant
to one item. Each translation equivalent received a median of 234
responses (\emph{Min} = 108, \emph{Max} = 880) from participants, both
languages pooled together.

@@

We developed the \emph{Exposure} predictor to account for the fact that
bilinguals' exposure to a given word-form is not only a function of the
word-form's lexical frequency, but also of the quantitative input they
receive from the language such word-form belongs to, we expressed
lexical frequencies as the product between both variables. First, we
extracted the child-directed lexical frequency of each word from the
CHILDES database
(\protect\hyperlink{ref-macwhinney2000childes}{MacWhinney, 2000};
\protect\hyperlink{ref-sanchez2019childesdb}{Sanchez et al., 2019}).
Using the corresponding lexical frequencies directly from Catalan and
Spanish was not possible due to the low number of Catalan participants
and tokens available in their corresponding CHILDES corpora, so they
were extracted from the English corpora instead. We mapped the lexical
frequencies of the English words to their Catalan and Spanish
translations (see \protect\hyperlink{ref-fourtassi2020growth}{Fourtassi
et al., 2020} for a similar approach), and transformed them to Zipf
scores (\protect\hyperlink{ref-vanheuven2014subtlexuk}{Van Heuven et
al., 2014}; \protect\hyperlink{ref-zipf1949human}{Zipf, 1949}). We
multiplied the resulting lexical frequencies by the reported degree of
exposure of the child to Catalan or Spanish. For instance, for a child
whose degree of exposure is 80\% for Catalan and 20\% for Spanish, the
expected \emph{Exposure} score to the Catalan word \emph{cotxe}
{[}\emph{car}{]}---with a lexical frequency of 6.33---would be 5.06,
while that of its translation to Spanish \emph{coche} would be 1.27.

We defined \emph{Cognateness} as the phonological similarity between
each word-form and its translation. For each translation equivalent, we
used the \texttt{stringdist}
(\protect\hyperlink{ref-vanderloo2014stringdist}{Loo, 2014}) R package
to calculate the Levenshtein distance between the Catalan and the
Spanish phonological transcriptions of the word-forms. The Levenshtein
distance measures the number of editions (insertions, deletions, or
substitutions) that one string of phonemes/characters must go through to
become identical to the other
(\protect\hyperlink{ref-levenshtein1966binary}{Levenshtein, 1966}). We
divided the Levenshtein distance of each translation equivalent by the
length of the longest word-form to correct for word length (longer
strings are likely to show a larger number of mismatches). Finally, we
subtracted the result from one so that it could be interpreted in terms
of phonological similarity, instead of phonological distance. This led
to a distance metric that ranged from zero to one, where zero indicates
that both word-forms are completely different (e.g.,
/\textipa{"taw.l5}/--/\textipa{"me.sa}/, \emph{table}), and one
indicates that the two word-forms are identical (e.g.,
/\textipa{"mar}/--/\textipa{"mar}/, \emph{sea}) (see
\protect\hyperlink{ref-floccia2018introduction}{Floccia et al., 2018};
\protect\hyperlink{ref-fourtassi2020growth}{Fourtassi et al., 2020};
\protect\hyperlink{ref-heeringa2003norwegian}{Heeringa \& Gooskens,
2003}; \protect\hyperlink{ref-laing2022phonological}{Laing, 2022} for
similar approaches;
\protect\hyperlink{ref-schepens2012distributions}{Schepens et al.,
2012}).

@@

We assessed the practical relevance of the estimated regression
coefficients of the model model following Kruschke \& Liddell
(\protect\hyperlink{ref-kruschke2018bayesian}{2018}). First, we
specified a region of practical equivalence (ROPE) from -0.025 to
+0.025, in the probability scale. This region indicates the range of
values that we considered equivalent to zero. We then summarised the
posterior distribution of each regression coefficient with the 95\%
highest density interval (HDI). This interval contains the true value of
this coefficient with 95\% probability, given the data. Finally, we
calculated the proportion of posterior samples in the 95\% HDI that fell
into the ROPE, noted as \(p(\text{ROPE})\), which indicates the
probability that the true value of the regression coefficient falls into
the ROPE (and therefore should be considered equivalent to zero). For
example, \(p(\text{ROPE})=.80\) indicates that, given our data, there is
a 80\% probability that the true value of the coefficient falls within
the ROPE, and can therefore be considered equivalent to zero.

We implemented the model using \texttt{brms}
(\protect\hyperlink{ref-burkner2017brms}{Bürkner, 2017}), a R interface
to the Stan probabilistic language (2.32.1)
(\protect\hyperlink{ref-carpenter2017stan}{Carpenter et al., 2017}). We
ran four iteration chains using the by-default No U-Turn Sampler
algorithm with 1,000 iterations each and an additional 1,000 warm-up
iterations per chain. Model posterior draws and predictions were handled
using the \texttt{tidybayes}
(\protect\hyperlink{ref-kay2021tidybayes}{Kay, 2021}) R package.
